---
title: "STAT2401 Analysis of Experiments "
author: "Final Examination Semester 1 2024"
date: "Monday 10th June 2024"
output:
  pdf_document: default
  html_notebook:
    highlight: haddock
    theme: flatly
  html_document:
    highlight: haddock
    theme: flatly
  word_document:
    highlight: haddock
---

# STAT2401 Final Examination - Semester 1 2024 (60 marks)

## Student name: Himakar Gadham

## Student ID:23783777


```{r echo=FALSE}
knitr::opts_chunk$set(prompt=FALSE, comment=NA, message=FALSE, tidy=TRUE, error=TRUE, warning=FALSE)
```

# Instructions

1. You are provided with Rmd template for your working that you are strongly recommended to submit under Question 1 File Response for partial marks. 

2. Save the RMD template (within the LMS, *right click ON the file, then Save Link As*) to your working directory and rename it in the format of *studentID_Surname.Rmd*, eg, *123456_Nur.Rmd*.

3. Marks for each question are shown. The total number of marks is 60.
  
    - This is a 2-hour examination.
    
    - You are provided with 15 minutes extra time for Rmarkdown template download, save and upload time.
    
    - There are 20 questions (Q2-Q21) set as multiple choice, multiple blanks, matching, calculated numeric. You must enter your answers on the LMS.
    
    - Include relevant R-output in your answers for some questions using R, if required.
    
         - You must provide your R code into the template provided here for full marks. 
         
         - Incorrect answers with R code attempt will be allocated some partial marks.
         
         - If you have any question or comment about the exam or would like to alert your Unit Coordinator to a perceived error, include a comment in your working in Rmd file, if appropriate, to indicate how you interpreted the question.
               
    - Carry out all of your work in this Rmarkdown file. 

4. **Save your work frequently!**

5. When you are finished, upload this *Rmarkdown file* back into *Question 1 File Response* on the LMS.

    -  You only need to submit this RMD file.
    
    -  You don't need to knit this file. 
    
    - DO NOT BE LATE. You must submit within the examination time (the last 15 minutes of 135 minutes). 
    
    - **We do not accept any Rmd submission by email.**


## Remarks

- Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

- You can add your working in this file whenever applicable, eg theoretical question. Clearly indicates the question number before your answer. 

- Make sure your program works for every chunk.

- It is your responsibility to call the libraries of R packages.

- The following chunk listed some R packages for the datasets that you may need to load as libraries.

```{r include=FALSE}
## Libraries from the computer labs
# library(tidyverse)
library(ggplot2)
library(PerformanceAnalytics)
library(leaps)
```


```{r}
## R packages for the datasets
library(HSAUR2)
library(MASS)
library(ISLR)
```


# Based on Questions 2-21 on the LMS. 

## Question 3

```{r}
library(MASS)
data(epil)
head(epil)
str(epil)
```

```{r}
progabide_data <- subset(epil, trt == "progabide")

kendall_result <- cor.test(progabide_data$y, progabide_data$age, method = "kendall")

kendall_result
test_statistic <- kendall_result$statistic
test_statistic
```



## Question 5

```{r}
Score<-c(4.58,4.36,4.19,4.57,4.39,4.52,4.37,4.55,4.62,4.65,4.63,4.52,4.43,4.71,4.73,4.76,4.52,4.71,4.52,4.64,4.86,4.66,4.65,4.84,5.02,4.78,4.65,4.81,4.96,4.79,5.02,4.84,4.62,4.84,5.00,5.11,5.08)
Years<-c(18,2,2,6,4,4,6,4,6,8,7,7,7,12,8,28,14,13,6,17,11,16,6,8,12,14,24,7,24,25,15,6,7,16,11,22,13)
Sales<-c(3.81,4.70,4.44,5.08,4.79,4.94,5.01,4.79,5.91,5.99,6.46,4.94,6.06,6.46,6.52,5.83,5.16,6.15,5.89,6.71,6.71,5.94,6.49,6.46,6.62,7.38,7.24,6.63,6.75,5.89,7.09,7.17,6.87,7.02,7.17,7.86,8.07)
Profit<-c(1.79,1.95,2.08,2.48,2.48,2.56,3.04,3.26,3.81,3.83,4.09,4.09,4.09,4.25,4.28,4.44,4.51,4.70,4.70,4.83,4.87,4.91,5.14,5.19,5.39,5.48,5.48,5.58,5.63,5.81,5.83,5.89,5.89,6.11,6.40,6.42,6.58)

credit.df=data.frame(Score,Years,Sales, Profit)
head(credit.df)

model <- lm(Score ~ Profit, data = credit.df)
summary(model)
r = 0.8123^2
r
```


## Question 6
```{r}
University<-c(1:33)
Year2008<-c(7018,7002,6926,6333,5432,5136,4994,4877,4726,4715,4405,4383,4335,4072,4047,3970,3872,3848,3819,3791,3761,3698,3575,3450,3204,3188,3132,2768,2725,2348,2256,3308,3374)
Year2014<-c(13706,13642,11738,11540,10690,8005,12106,10634,7386,6385,8231,8679,9300,9497,7656,8046,7750,8062,6435,7569,6802,8310,8532,6584,6544,7278,5524,5397,7042,5542,3256,6040,7844)
tuition.df <-data.frame(University,Year2008,Year2014)
head(tuition.df)
model <- lm(Year2014 ~ Year2008, data = tuition.df)
summary(model)
```
```{r}
conf_interval <- confint(model, level = 0.99)
conf_interval
```
```{r}
prediction_interval <- predict(model, newdata = data.frame(Year2008 = 5500), interval = "prediction", level = 0.90)
prediction_interval
```


## Question 8
```{r}
Score<-c(4.58,4.36,4.19,4.57,4.39,4.52,4.37,4.55,4.62,4.65,4.63,4.52,4.43,4.71,4.73,4.76,4.52,4.71,4.52,4.64,4.86,4.66,4.65,4.84,5.02,4.78,4.65,4.81,4.96,4.79,5.02,4.84,4.62,4.84,5.00,5.11,5.08)
Years<-c(18,2,2,6,4,4,6,4,6,8,7,7,7,12,8,28,14,13,6,17,11,16,6,8,12,14,24,7,24,25,15,6,7,16,11,22,13)
Sales<-c(3.81,4.70,4.44,5.08,4.79,4.94,5.01,4.79,5.91,5.99,6.46,4.94,6.06,6.46,6.52,5.83,5.16,6.15,5.89,6.71,6.71,5.94,6.49,6.46,6.62,7.38,7.24,6.63,6.75,5.89,7.09,7.17,6.87,7.02,7.17,7.86,8.07)
Profit<-c(1.79,1.95,2.08,2.48,2.48,2.56,3.04,3.26,3.81,3.83,4.09,4.09,4.09,4.25,4.28,4.44,4.51,4.70,4.70,4.83,4.87,4.91,5.14,5.19,5.39,5.48,5.48,5.58,5.63,5.81,5.83,5.89,5.89,6.11,6.40,6.42,6.58)

credit.df=data.frame(Score,Years,Sales, Profit)
head(credit.df)

model1 <- lm(Score ~ Sales, data = credit.df)
summary(model1)
anova(model1)
```


## Question 9
```{r}
Score<-c(4.58,4.36,4.19,4.57,4.39,4.52,4.37,4.55,4.62,4.65,4.63,4.52,4.43,4.71,4.73,4.76,4.52,4.71,4.52,4.64,4.86,4.66,4.65,4.84,5.02,4.78,4.65,4.81,4.96,4.79,5.02,4.84,4.62,4.84,5.00,5.11,5.08)
Years<-c(18,2,2,6,4,4,6,4,6,8,7,7,7,12,8,28,14,13,6,17,11,16,6,8,12,14,24,7,24,25,15,6,7,16,11,22,13)
Sales<-c(3.81,4.70,4.44,5.08,4.79,4.94,5.01,4.79,5.91,5.99,6.46,4.94,6.06,6.46,6.52,5.83,5.16,6.15,5.89,6.71,6.71,5.94,6.49,6.46,6.62,7.38,7.24,6.63,6.75,5.89,7.09,7.17,6.87,7.02,7.17,7.86,8.07)
Profit<-c(1.79,1.95,2.08,2.48,2.48,2.56,3.04,3.26,3.81,3.83,4.09,4.09,4.09,4.25,4.28,4.44,4.51,4.70,4.70,4.83,4.87,4.91,5.14,5.19,5.39,5.48,5.48,5.58,5.63,5.81,5.83,5.89,5.89,6.11,6.40,6.42,6.58)

credit.df=data.frame(Score,Years,Sales, Profit)
head(credit.df)
model <- lm(Score ~ Profit, data = credit.df)
plot(model)

```
```{r}
leverage_values <- hatvalues(model)
n <- length(Score) 
p <- 1 
threshold <- 2 * (p + 1) / n
leverage_points <- which(leverage_values > threshold)
leverage_points
```



## Question 10

```{r}
Crews <-c(16,10,12,16,16,4,2,4,6,2,12,8,16,2,2,2,6,10,16,16,10,6,2,6,10,12,4,4,16,8,10,16,6,10,12,8,10,8,8,2,16,8,8,12,10,16,2,2,8,12,4,4,12)

Rooms <-c(51,37,37,46,45,11,6,19,29,14,47,37,60,6,11,10,19,33,46,69,41,19,6,27,35,55,15,18,72,22,55,65,26,52,55,33,38,23,38,10,65,31,33,47,42,78,6,6,40,39,9,22,41)
data <- data.frame(Crews, Rooms)

model_original <- lm(Rooms ~ Crews, data = data)
summary(model_original)
data$sqrt_Crews <- sqrt(data$Crews)
data$sqrt_Rooms <- sqrt(data$Rooms)

model_trans <- lm(sqrt_Rooms ~ sqrt_Crews, data = data)
summary(model_trans)

s_transformed <- sqrt(sum(residuals(model_trans)^2) / df.residual(model_trans))
s_transformed

plot(model_original)
```

## Question 11
```{r}
data("airquality") # this data available in R base package, you don't need to install any R package
View(airquality) # to view the dataset

head(airquality) # to view the first 6 rows

m1 <- lm(Ozone ~ Solar.R + Wind + Temp, data = airquality)
summary(m1)
anova_model <- anova(m1)
anova_model
```



## Question 13
```{r}
library(HSAUR2)
attach(USairpollution)

pollution <- USairpollution
#View(pollution) # 41 rows 
head(pollution)

correlation_matrix <- cor(pollution[, -1])
correlation_matrix

cor_max <- which(correlation_matrix == max(correlation_matrix[upper.tri(correlation_matrix)]), arr.ind = TRUE)
colnames(correlation_matrix)[cor_max]
```

```{r}
model <- lm(SO2 ~ temp + manu + popul + wind + precip + predays, data = pollution)
summary(model)

library(car)
vif_values <- vif(model)
vif_manu <- vif_values["manu"]
vif_manu
```

## Question 14

```{r}
Score<-c(4.58,4.36,4.19,4.57,4.39,4.52,4.37,4.55,4.62,4.65,4.63,4.52,4.43,4.71,4.73,4.76,4.52,4.71,4.52,4.64,4.86,4.66,4.65,4.84,5.02,4.78,4.65,4.81,4.96,4.79,5.02,4.84,4.62,4.84,5.00,5.11,5.08)
Years<-c(18,2,2,6,4,4,6,4,6,8,7,7,7,12,8,28,14,13,6,17,11,16,6,8,12,14,24,7,24,25,15,6,7,16,11,22,13)
Sales<-c(3.81,4.70,4.44,5.08,4.79,4.94,5.01,4.79,5.91,5.99,6.46,4.94,6.06,6.46,6.52,5.83,5.16,6.15,5.89,6.71,6.71,5.94,6.49,6.46,6.62,7.38,7.24,6.63,6.75,5.89,7.09,7.17,6.87,7.02,7.17,7.86,8.07)
Profit<-c(1.79,1.95,2.08,2.48,2.48,2.56,3.04,3.26,3.81,3.83,4.09,4.09,4.09,4.25,4.28,4.44,4.51,4.70,4.70,4.83,4.87,4.91,5.14,5.19,5.39,5.48,5.48,5.58,5.63,5.81,5.83,5.89,5.89,6.11,6.40,6.42,6.58)

credit.df=data.frame(Score,Years,Sales, Profit)
head(credit.df)
model <- lm(Score ~ Years + Sales, data = credit.df)
summary(model)

n <- nrow(credit.df)
p <- 2
n
p
```


## Question 15
```{r}
library(MASS) 
attach(Boston)

head(Boston) # to view the first 6 rows

model_reduced <- lm(medv ~ lstat + rm + ptratio, data = Boston)
model_full <- lm(medv ~ lstat + rm + ptratio + dis + nox + chas + black, data = Boston)
anova(model_reduced, model_full)
```


## Question 16
```{r}
library(ISLR) 

attach(Carseats)

#head(Carseats) # to view the first 6 rows

full <- regsubsets(Sales ~ ., data = Carseats, nvmax = 11)
AllSubsets.summary <- summary(full)
AllSubsets.summary

modelf <- lm(Sales ~ CompPrice+Advertising + Price + ShelveLoc, data = Carseats)
```

```{r}
par(mfrow = c(1, 3))
par(cex.axis = 1.5)
par(cex.lab = 1.5)

plot(1:11, AllSubsets.summary$adjr2, xlab = "subset size", ylab = "adjusted R-squared",
    type = "b")
plot(1:11, AllSubsets.summary$cp, xlab = "subset size", ylab = "Mallows' Cp", type = "b")
abline(0, 1, col = 2)
plot(1:11, AllSubsets.summary$bic, xlab = "subset size", ylab = "BIC", type = "b")
adj <-AllSubsets.summary$adjr2[7]
adj
```
## Question 17
```{r}
library(ISLR) 

attach(Carseats)

#head(Carseats) # to view the first 6 rows

full_model <- lm(Sales ~ ., data = Carseats)
summary(full_model)
AIC(full_model)
```

```{r}
backward_model <- step(full_model, direction = "backward", trace = 0)
summary_backward_model <- summary(backward_model)
summary_backward_model
sigma <- summary_backward_model$sigma
sigma
```


## Question 18
```{r}
library(ISLR)

data(College)
head(College)
str(College)

null_model <- lm(Apps ~ 1, data = College)
full_model <- lm(Apps ~ ., data = College)
```
```{r}
forward_model <- step(null_model, scope = list(lower = null_model, upper = full_model), direction = "forward", trace = 0)
summary_forward_model <- summary(forward_model)
summary_forward_model
r_squared <- summary_forward_model$r.squared
r_squared
```


## Question 19

```{r}
library(ISLR) 

attach(Carseats)

head(Carseats) # to view the first 6 rows

null_model <- lm(Sales ~ 1, data = Carseats)
full_model <- lm(Sales ~ ., data = Carseats)

forward_model <- step(null_model, scope = list(lower = null_model, upper = full_model), direction = "forward")
summary_forward_model <- summary(forward_model)
summary_forward_model

library(DAAG)
press(forward_model)
```


## Question 20
```{r}
plot(forward_model)
```

```{r}
str(Carseats)
standardized_residuals <- rstandard(forward_model)
plot(standardized_residuals)
abline(h = c(-3, 3), col = "red", lty = 2)

outliers <- which(standardized_residuals < -3 | standardized_residuals > 3)
outliers
length(outliers)
```


## Question 21
```{r}
library(ISLR)

data(College)
head(College)
str(College)

M1 <- lm(Apps ~ Accept * Private, data = College)

M2 <- lm(Apps ~ Accept + Private, data = College)

anova(M2, M1)
```

